Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2020-11-22 manu_draft.Rmd
started discussion

We present a simple visual comparison of chlorophyll estimates from models SY, SYD, and SYDI to emphasize that similar fits can be achieved by all of the presented models (Figure \@ref(fig:modsumfig), S is identical to SY and is not shown).  Models SY, SYD, and SYDI were fit to chlorophyll data from station 34 using large k values for the arguments `num_knots_y`, `num_knots_D`, `num_knots_Y_ti`, and `num_knots_D_ti` for each model.  Model predictions by day of year are visually similar between the models (Figure \@ref(fig:modsumfig)a) and are very similar when compared to a 1:1 line (Figure \@ref(fig:modsumfig)b).  However, when comparing similarities among the models using only the continuous year smoother (`s(cont_year)`, as an eexample), the models differ substantially (Figure \@ref(fig:modsumfig)) because of structural differences and differences in the penalties applied to te basis functions.  These results are also demonstrated in differences between the effective degrees of freedom among the additive components of each model in Table \@ref(tab:modstrtab). Accordingly, the models may differ in which structural component describes variation in the chlorophyll time series but the model predictions are very similar.

@@

@@

Explain how choosing window widths and time periods is a user choice that is arbitrary, identifying an objective approach to identifying trends can pave the way for future work (parallels with mgcv?)

Benefits of mixed-meta to provide accurate results, reduces possibility of type I/II errors (e.g., a/b examples in trndcmpex)

>>>>

We present a simple visual comparison of chlorophyll estimates from models SY, SYD, and SYDI to emphasize that similar fits can be achieved by all of the presented models (Figure \@ref(fig:modsumfig), S is identical to SY and is not shown).  Models SY, SYD, and SYDI were fit to chlorophyll data from station 34 using large k values for the arguments `num_knots_y`, `num_knots_D`, `num_knots_Y_ti`, and `num_knots_D_ti` for each model.  Model predictions by day of year are visually similar between the models (Figure \@ref(fig:modsumfig)a) and are very similar when compared to a 1:1 line (Figure \@ref(fig:modsumfig)b).  However, when comparing similarities among the models using only the continuous year smoother (`s(cont_year)`, as an example), the models differ substantially (Figure \@ref(fig:modsumfig)) because of structural differences and differences in the penalties applied to the basis functions.  These results are also demonstrated in differences between the effective degrees of freedom among the additive components of each model in Table \@ref(tab:modstrtab). Accordingly, the models may differ in which structural component describes variation in the chlorophyll time series but the model predictions are very similar.

@@

The approach developed here can be used to estimate seasonal averages with uncertainty in water quality parameters and evaluate trends in these seasonal changes across years using full propagation of error estimates across methods.  This approach leverages the ability of GAMs to objectively estimate smoothed trends across years by identifying an "optimal" level of smoothing with the mgcv package that uses generalized cross-validation to extract an underlying signal in the observed data. Importantly, specifying an upper limit for the number of knots that can potentially be used to fit the smoother is critical to this approach.  A smaller limit can lead to under-smoothing and an insufficient characterization of trends that risk inflated Type I error rates. We further show that models with different structures in the smoothed components, such as separate smoothers for continuous year and day of year, can produce nearly identical results in the predicted trends if the knots are sufficiently high to allow the GAMs to be fit as intended.  This comparison demonstrated that model structure was of less importance than allowing the model sufficient freedom to estimate the trends over time. Although models with different structures allocate variance components of the response variable differently, this is a trivial issue for the second and third stage of analyses developed herein.

The extension of the GAM results to assess trends at an individual station is a new approach that previous trend analysis methods cannot sufficiently address.  Although the development of our approach was motivated by specific questions regarding assessment of seasonal changes over time, the application has advantages over more conventional methods that inadequately account for time series characteristics of water quality data from long-term monitoring programs. In particular, missing observations or irregular sampling can complicate trend assessment and comparison of trends between locations that may differ in sampling design.  As noted above, non-parametric approaches (i.e., seasonal Kendall tests) are by far the most common trend analysis methods applied to long-term water quality data.  These methods only assess the direction and significance of comparisons between year pairs, and importantly, do not account for full propagation of uncertainty inherent in raw observations.  Aggregation of raw data, e.g., averaging of observations within a year or season to comply with the requirements of Kendall tests, risks loss of information by removing variation between observations at smaller time scales.  The logical outcome is increased likelihood of incorrect conclusions from test results.

Our example in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate how different conclusions can be obtained if propagation of uncertainy from raw observations across methods is unaccounted for in trend assessment.  These comparisons are analogous to the example in the previous paragraph where a common approach prior to trend analysis is to aggregate raw observations.  Our assessment of trends using linear regression applied to seasonal averages from the raw observations is effectively similar to averaging results within a year and applying Kendall test. In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data.  This is particularly problematic if sampling effort varies substantially between years.  Averages may be skewed in a particular direction if an annual estimates are based on a handful of observations from select months (e.g., summer only).  The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages.  Further and most importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-meta regression analyses.  As a result, trend assessments from mixed-meta regression should be considered an accurate representation of the actual trend within the limits of the raw observations.

The underlying cross-validation methods used by GAMs in the mgcv package also reduce the decisions required by an analyst that may be necessary for the implementation of alternative trend assessment methods.  For example, WRTDS and similar smoothing approaches (e.g., LOESS) require decisions on appropriate window widths or bandwidths to define the neighborhood of observations for smoothing.  Until implementation of generalized cross-validation in GAMs, these decisions were also required, leading to potential under- or over-smoothing of fitted results to the raw observations.  Because these decisions are no longer needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are not subject ot process or observation error.  However, we acknowledge that the third stage analyses that require explicit user inputs on year periods to define trends partially impose arbitrary decisions that can influence interpretation of results.  Although there are undoubtedly many scenarios where years of interest can be chosen obectively (e.g., regulatory compliance periods), a more general question of "when" changes occur independent of user decisions is also important to address.  For example, trends could be assessed based on five or ten year moving windows, but which result should be used to inform decisions?  There may be additional methods to develop using objective criteria to more accurately identify inflection points or other important periods where changes occur independent of a user choice.

The trend assessment approach also does not address issues of causality in the observed trends.  In this regard, the approach is similar to other trend assessment methods where the focus is understanding direction, magnitude, and confidence in changes in water quality variables over time. However, a logical follow-up question is "what" is driving the trend after the trend has been adequately described.  This information has obvious implications for management decisions on factors that influence water quality changes, e.g., wastewater treatment upgrades, large-scale climatic factors, or flow regulation practices.  An advantage of GAMs is their flexibility in including alternative predictors, such that the significance of a predictor or comparison of nested models with and without different predictors can provide evidence of which predictors are driving the observed trends. In such cases, considerations of model structure have direct implications on conclusions regarding potential causality.  As noted above, model structure in describing the long-term trend component was irrelevant for describing long-term trends, although a distinction here is made in the objective of the analysis.  Our goal herein was to describe chlorophyll changes relative to time where the predictors were variations on a general theme (e.g., season vs. year), whereas using GAMs with different predictors assess potential causality is a different application.  Therefore, using our approach to assess causality will require considerations of model structure given how GAMs could be used to asssess different questions.

@@

Application to other water quality variables and opportunity for ecological interpretation (cue up next paper?)
